---
title: "RCX - an R package implementing the Cytoscape Exchange (CX) format"
author: 
- name: Florian J. Auer
  email: Florian.Auer@med.uni-goettingen.de
  affiliation: &id University Medical Center Göttingen
- name: Zaynab Hammoud
  email: Zaynab.Hammoud@med.uni-goettingen.de
  affiliation: *id
- name: Frank Kramer
  email: Frank.Kramer@med.uni-goettingen.de
  affiliation: *id
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RCX Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Networks are a powerful and flexible methodology for expressing
biological knowledge for computation and communication. Albeit 
its benefits, the sharing of networks, the collaboration on 
network curation, keeping track of changes between 
different network versions, and detecting different versions itself,
still is a major problem in network biology.

The Network Data Exchange, or NDEx, is an open-source software framework 
to manipulate, store and exchange networks of various types and formats (Pratt et al., 2015, Cell Systems 1, 302-305, October 28, 2015 ©2015 Elsevier Inc. [ScienceDirect](http://www.sciencedirect.com/science/article/pii/S2405471215001477)). 
NDEx can be used to upload, share and publicly distribute networks, while providing
an output in formats, that can be used by plenty of other applications.

The public NDEx server is a network data commons which provides pathway collections like the Pathway Interaction Database of the NCI (http://www.ndexbio.org/#/user/301a91c6-a37b-11e4-bda0-000c29202374) and the Cancer Cell Maps Initiative (http://www.ndexbio.org/#/user/b47268a6-8112-11e6-b0a6-06603eb7f303). 

This package provides an interface to query the public NDEx server, as well as private installations, in order to upload, download or modify biological networks. 

This document aims to help the user to install and benefit from the
wide range of funtionality of this implementation.
The package also provides classes to implement the Cytoscape
Cyberinfrastructure (CX) Format and to extend the [iGraph Package] (http://igraph.org/r/).

The package is compatible with both NDEx versions 1.3 and 2.0.

# Installation
## Installation via GitHub
using [*devtools*](http://cran.r-project.org/web/packages/devtools/index.html) R package
```{r, eval=FALSE}
require(devtools)
install_github("frankkramer-lab/RCX")
library(RCX)
```

# Quick Start

ToDo!



# RCX

For the exchange of network data, NDEx uses the Cytoscape
Cyberinfrastructure Network Interchange Format, or just CX format (See
[*http://www.home.ndexbio.org/data-model/*](http://www.home.ndexbio.org/data-model/)).
CX is an Aspect-Oriented Network Interchange Format encoded in JSON,
which is used as basis for the R implementation of the CX format, namely
RCX.

The RCX object is currently implemented within this package as a list of
data.frames, containing meta-data and all aspects of the network. The
structure of an RCX object, as shown via `str(rcx)` could be a list like
this:

```{r}
library(RCX)
rcx = rcx_new(data.frame('@id'=c(1,2,3),n=c('Some Name','And another name',NA), r=c('HGNC:Symbol',NA,'UniProt:C3P0'), check.names=FALSE))

str(rcx, max.level = 2)
```

The data.frames representing nodes and edges could look like this:

```{r}
rcx[["nodes"]][1:5,]

rcx[["edges"]][1:5,]
```

Usually, and RCX object is automatically created by using the functions
of this package for downloading network data from a NDEx server. But it
might be useful to convert an RCX object from/to JSON manually, for
example for down-/uploading a CX file from/to a NDEx server via the web
interface. For handling the network information within R, besides RCX
objects, one can use RCXgraph objects. A lossless conversion between the
two files can be done using the following functions:

```{r, warning=FALSE}
## convert RCX to JSON
json <- rcx_toJSON(rcx)

## ...and back
rcx <- rcx_fromJSON(json)

## convert RCX to RCXgraph
rcxgraph <- rcx_toRCXgraph(rcx)

## ...and back
rcx <- rcxgraph_toRCX(rcxgraph)
```

It is possible to create blank RCX objects from scratch:

```{r}
newRcx <- rcx_new()
```

After a RCX object is downloaded from an NDEx server, it will contain
some aspects that are not present in a newly generated network, i.e.
“ndexStatus”, “provenanceHistory” and “status”. Removing those aspects
might be useful in some cases.

```{r}
asNewRcx <- rcx_asNewNetwork(rcx)
```

After a RCX object underwent some changes (adding/removing
nodes/edges/aspects/meta- data, etc.), it is advisable to check a RCX
object for its integrity and update its meta-data.

```{r,eval=F}
rcx <- rcx_updateMetaData(rcx)
```

The meta-data is not only updated by the function, but also created, if
not existent in the first place. It is necessary to mention, that the
function tries to check the format and content of most of the aspects
and properties, but due to the dynamic structure of RCX and CX, it is
advised to have a look at the CX data model specification for a deeper
insight about the core structure, dependencies and limitations.
