---
title: "01. RCX - an R package implementing the Cytoscape Exchange (CX) format"
author: 
- name: Florian J. Auer
  email: Florian.Auer@informatik.uni-augsburg.de
  affiliation: &id Augsburg of University
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. RCX - an R package implementing the Cytoscape Exchange (CX) format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Networks are a powerful and flexible methodology for expressing
biological knowledge for computation and communication. Albeit 
its benefits, the sharing of networks, the collaboration on 
network curation, keeping track of changes between 
different network versions, and detecting different versions itself,
still is a major problem in network biology.

The Network Data Exchange, or NDEx, is an open-source software framework 
to manipulate, store and exchange networks of various types and formats (Pratt et al., 2015, Cell Systems 1, 302-305, October 28, 2015 ©2015 Elsevier Inc. [ScienceDirect](http://www.sciencedirect.com/science/article/pii/S2405471215001477)). 
NDEx can be used to upload, share and publicly distribute networks, while providing
an output in formats, that can be used by plenty of other applications.

The public NDEx server is a network data commons which provides pathway collections like the Pathway Interaction Database of the NCI (http://www.ndexbio.org/#/user/301a91c6-a37b-11e4-bda0-000c29202374) and the Cancer Cell Maps Initiative (http://www.ndexbio.org/#/user/b47268a6-8112-11e6-b0a6-06603eb7f303). 

CX is a JSON based data structure designed as flexible model for the exchange and transmission of networks.
Its key-features are flexibility, modularity, and extensibility, with the focus on transmitting the network data. 
Although those features are widely used in common REST protocols they don't quite fit to the R way of thinking about data.

This package provides an adaption of the CX format to R data formats and types, in order to create and modify, load and export, and visualize networks.
This document aims to help the user to install and benefit from the wide range of functionality of this implementation.

# Installation
## Installation via GitHub
using [*devtools*](http://cran.r-project.org/web/packages/devtools/index.html) R package
```{r, eval=FALSE}
require(devtools)
install_github("frankkramer-lab/RCX")
library(RCX)
```


# The Cytoscape Exchange (CX) Format 

CX is an "aspect-oriented" data structure, which means that the network is divided into several independent modules ("aspects").
Each aspect has its own schema for its contained information, and links between aspects are realized by referencing the unique internal IDs of other aspects.

A simple example is a network only containing nodes and edges, where both are individual aspects and the edges are defined by referencing the IDs of the nodes as sources and targets.

```
nodes: 0:A, 1:B, 2:C, 3:D, 4:E
edges: 0--1, 0--2, 1--3, 2--4
```

The CX format comes with some pre-defined aspects, which are divided in two groups:
- core aspects, with which the most properties of the network can be defined
- cytoscape aspects, which provide additional information to visualize the networks with [Cytoscape](https://cytoscape.org/)

Additionally, it is possible to extend the CX format by defining own aspects if needed.

In the following, it will be explained, how to read and write networks from/to CX files, create and modify RCX networks, validate its contents and finally visualize them.

# The basics

```{r, echo=FALSE, message=FALSE}
library(RCX)
```

## Read and write CX files

Networks can be downloaded from the [NDEx plattform](http://ndexbio.org/) as files in JSON format.
Those files can be read as following:

```{r readRCX}
cxFile = system.file(
  "extdata", 
  "Imatinib-Inhibition-of-BCR-ABL-66a902f5-2022-11e9-bb6a-0ac135e8bacf.cx", 
  package = "RCX"
)
rcx = readRCX(cxFile)
```

To check the contents of the RCX object, a summary of the object can be shown:

```{r summaryCX, results='hide'}
summary(rcx)
```

RCX objects can be saved in a similar manner:

```{r writeCX, eval=FALSE}
writeCX(rcx, "some-file.cx")
```

However, there might some errors occur while reading CX file.
This might happen, because the definition of the CX has changed over time, and so the definition of some aspects. 
Therefore it is possible, that there are still some networks stored at the NDEx plattform following a deprecated format.
In those cases it might be helpful to process the CX network step by step:

- just read the JSON without parsing
```{r readJSON}
json = readJSON(cxFile)
```
- parse the JSON
```{r parseJSON}
aspectList = parseJSON(json)
```
- process the aspect data
```{r processCX}
rcx = processCX(aspectList)
```

This also allows to handle CX networks in JSON format, even if it comes from a different source instead of a file.

All the above function for processing CX networks come with an option to show the performed steps.
This might be helpful for finding ocurring errors:

```{r readRCXverbose}
rcx = readRCX(cxFile, verbose = TRUE)
```


# Core aspects




# RCX

For the exchange of network data, NDEx uses the Cytoscape
Cyberinfrastructure Network Interchange Format, or just CX format (See
[*http://www.home.ndexbio.org/data-model/*](http://www.home.ndexbio.org/data-model/)).
CX is an Aspect-Oriented Network Interchange Format encoded in JSON,
which is used as basis for the R implementation of the CX format, namely
RCX.

The RCX object is currently implemented within this package as a list of
data.frames, containing meta-data and all aspects of the network. The
structure of an RCX object, as shown via `str(rcx)` could be a list like
this:

```{r, eval=FALSE}
library(RCX)
rcx = rcx_new(data.frame('@id'=c(1,2,3),n=c('Some Name','And another name',NA), r=c('HGNC:Symbol',NA,'UniProt:C3P0'), check.names=FALSE))

str(rcx, max.level = 2)
```

The data.frames representing nodes and edges could look like this:

```{r, eval=FALSE}
rcx[["nodes"]][1:5,]

rcx[["edges"]][1:5,]
```

Usually, and RCX object is automatically created by using the functions
of this package for downloading network data from a NDEx server. But it
might be useful to convert an RCX object from/to JSON manually, for
example for down-/uploading a CX file from/to a NDEx server via the web
interface. For handling the network information within R, besides RCX
objects, one can use RCXgraph objects. A lossless conversion between the
two files can be done using the following functions:

```{r, warning=FALSE, eval=FALSE}
## convert RCX to JSON
json <- rcx_toJSON(rcx)

## ...and back
rcx <- rcx_fromJSON(json)

## convert RCX to RCXgraph
rcxgraph <- rcx_toRCXgraph(rcx)

## ...and back
rcx <- rcxgraph_toRCX(rcxgraph)
```

It is possible to create blank RCX objects from scratch:

```{r, eval=FALSE}
newRcx <- rcx_new()
```

After a RCX object is downloaded from an NDEx server, it will contain
some aspects that are not present in a newly generated network, i.e.
“ndexStatus”, “provenanceHistory” and “status”. Removing those aspects
might be useful in some cases.

```{r, eval=FALSE}
asNewRcx <- rcx_asNewNetwork(rcx)
```

After a RCX object underwent some changes (adding/removing
nodes/edges/aspects/meta- data, etc.), it is advisable to check a RCX
object for its integrity and update its meta-data.

```{r,eval=F}
rcx <- rcx_updateMetaData(rcx)
```

The meta-data is not only updated by the function, but also created, if
not existent in the first place. It is necessary to mention, that the
function tries to check the format and content of most of the aspects
and properties, but due to the dynamic structure of RCX and CX, it is
advised to have a look at the CX data model specification for a deeper
insight about the core structure, dependencies and limitations.
